---
title: "Intro2dplyr"
author: "Justin Strate"
output: html_document
---

```{r}
library(dplyr)
data("mtcars")

head(mtcars)
str(mtcars)

data("iris")

head(iris)
str(iris)

class(mtcars)
class(iris)
mtcars
iris
```

dplyr has the function tbl_df which we can use to convert data frames into tbl_df's

```{r}
mtcars <- tbl_df(mtcars)
iris <- tbl_df(iris)
```

Let's save the names of the cars
```{r}
vehicle <- rownames(mtcars)
```
tbl_df's print differently and they run faster

```{r}
mtcars
iris
```

If you want to print more rows you can use the n argument in the print function
```{r}
print(iris, n=52)
```

5 basic function you want to keep in mind when looking at dplyr
select, filter, arrange, mutate, and summarise
summarise is espeically useful with the function group_by
also covers briefly the tbl_df, rename, everything


#1.) select
select takes after sql and serves to select variables of interest from a certain data frame
for example let's say we want to select wt and mpg from the mtcars data set
```{r}
select(mtcars, wt, mpg)
```

The base R approach to doing this is as follows
```{r}
mtcars[,c('wt', 'mpg')]
```

You can also use pipelining
the %>% is supposed to be interpreted as "then"
let's also rename the variables using the rename function
```{r}
mtcars %>%
  select(mpg,wt, gear) %>%
  rename(Weight = wt, 'Miles Per Gallon' = mpg)
```

Notice the order in which you selects the variables reflects how they are outputted.

Sometimes, you may simply want to rearrange only on and leave all the other columns.
For example say you want to rearrange iris so that Species is the first column in the data frame. One approach is to use the \code{everything()} function.
```{r}
select(iris, Species, everything())
```

You can give select a range of variables using the colon. For example suppose we want to select all the variables from wt to mpg.
```{r}
select(mtcars, wt:mpg)
```
We can select all the variables in the data set except for a few variables as follows
```{r}
select(mtcars, -wt, -mpg)
```

base R approach
```{r}
mtcars[,!names(mtcars) %in% c('wt', 'mpg')]
```

#filter
The next verb is filter, which as the name implies filters the data based on a condition. For example let's say we only want v-engine vehicles
```{r}
filter(mtcars, vs==1)
```

The base R approach to this is to use
```{r}
mtcars[mtcars$vs==1,]
```

We can also filter on more than one condition for example let's say we want the v engines and 4 cylinder.
```{r}
filter(mtcars, vs==1 & cyl==4)
```

The base R approach
```{r}
mtcars[mtcars$vs==1 & mtcars$cyl==4,]
```
We can drop vs and cyl since we know they can only be 1 and 4 respectively
```{r}
mtcars %>%
  filter(vs==1 & cyl==4) %>%
  select(-vs, -wt)
```

The base R approach
```{r}
mtcars[mtcars$vs==1 & mtcars$cyl==4, !colnames(mtcars) %in% c('mpg', 'cyl')]
```

#mutate- add in a variable
First let me create a copy of mtcars
```{r}
mtcars_copy <- mtcars
```

Let's create a new variable in the data frame called- mpg2wt which mpg/wt
```{r}
mtcars <- mtcars %>%
  mutate(mpg2wt = mpg/wt)
```
The base R approach
```{r}
mtcars[,'mpg2wt'] <- mtcars$mpg/mtcars$wt
mtcars
```
You can add more than one variable at a time.
For example let's add and let's add the variable vehicle which we saved earlier
Let's also rearrange the data using what we learned earlier so that vehicle
```{r}
mtcars <- mtcars %>%
  mutate(hp2wt = hp/wt, vehicle = vehicle) %>%
  select(vehicle, everything())
mtcars
```

#arrange- 
arranges data in a certain order (ascending by default for numeric or a-z for character)
```{r}
mtcars %>%
  select(vehicle, mpg, wt) %>%
  arrange(mpg)
```
The base R approach
```{r}
mtcars[order(mtcars$mpg),names(mtcars) %in% c('vehicle', 'mpg', 'wt')]
```
If you want the data in descending order you can use the desc function inside arrange
```{r}
mtcars %>%
  select(vehicle, mpg, wt) %>%
  arrange(desc(mpg))
```

We can also arrange by a character varable if you want the data in alphabetical order
```{r}
mtcars %>%
  select(vehicle, mpg, wt) %>%
  arrange(vehicle)
```

desc also works with character variables if you want them to go in descending alphabetical order
```{r}
mtcars %>%
  select(vehicle, mpg, wt) %>%
  arrange(desc(vehicle))
```
You can also arrange by more than one variable. For example suppose we want to arrange the data by ascending cyl but descending qsec and let's print 15 observations
```{r}
mtcars %>%
  select(vehicle, cyl, qsec) %>%
  arrange(cyl, desc(qsec)) %>%
  print(n=15)
```

#5.) summarise
summarise helps us summarise data
For example, let's say we want the the mean and standard deviations of mpg and wt
```{r}
mtcars%>%
  select(mpg, wt) %>%
  summarise(mean_mpg = mean(mpg), sd_mpg = sd(mpg), mean_wt = mean(wt), sd_wt = sd(wt))
```


summarise is espcially useful in conjunction with the function group_by.
For example, let's say we want the mean sepal length for each different species in the iris data set
```{r}
iris %>%
  select(Species, Sepal.Length) %>%
  group_by(Species) %>%
  summarise(mean_sepal_length = mean(Sepal.Length))
```
The base R approach
```{r}
tapply(iris$Sepal.Length, iris$Species, mean)
```
Let's say we want to know how many of each species we have- 50 as you can tell from earlier
We can use the n function
```{r}
iris %>%
  select(Species) %>%
  group_by(Species) %>%
  summarise(Count=n())
```
The base R approach
```{r}
tapply(iris$Sepal.Length, iris$Species, length)
```
You can also use this with more than 1 variable
for example let's see how many v- engines and straigth engines (vs=0) we have for each number of cylinder
```{r}
mtcars %>%
  select(vehicle, cyl, vs) %>%
  group_by(cyl, vs) %>%
  summarise(Count = n())
```
This should be interpreted as there is 1 4 cylinder engine that is also a straight engine (vs=0)
There are 10 v 4 cylinder engines and so forth
There are no 8 cylinder v engines

The base R approach
```{r}
with(mtcars, tapply(mpg, list(cyl, vs), length))
```