---
title: "More Advanced Data Wrangling"
author: "Justin Strate"
date: "January 16, 2018"
output: html_document
---

```{r}
library(dplyr)
library(tidyr)
```

## Introduction
The tidyr package allows you to easily transform your data in ways that would otherwise be very difficult with base R. The four main functions I would like to introduce you to are `gather`, `seperate`, `collapse`, and `spread`. The first data set I would like to use the iris data set. It contains 3 species of flowers: setosa, versicolor, and virginica. The are 50 observations of sepal length, sepal width, petal length, and pedal width; altogether, 150 observations.

```{r}
data("iris")
head(iris)
```

Imagine we wanted a data frame with one 3 columns: one the species, the type of measurement (sepal length, sepal width, etc) and the recorded. We coud use the `gather` function for this. Following in the tidyverse ways, the first argument in the gather function is the data set, which we will supply using the magrittr pipe (%>%). The second argument, key, is the name of the new column that will indicate the original column column name. In our case we will name this column "measurement_type", and it will read either "Sepal.Length", "Sepal.Width", "Pedal.Length" or "Pedal.Width". The third argument, value, is the name of the new column being created; in our case we will just name it "new_col". The last argument is the columns you are gathering. These can be suppled column by column seperated by a comma or you can specify a range of columns using the ":" operator as I do here.

```{r}
iris2 <- iris %>%
  gather( key = "measurement_type", value = "new_col", Sepal.Length:Petal.Width)

head(iris2)
```


I have run into scenarios where I had a data set indexed by weeks, but I wanted a data set that was indexed by days. I've created one such data set below.

```{r}
dat <- tbl_df(data.frame(
  week_begins = as.Date(c('12/31/17', '01/07/18', '01/14/18'), format = "%m/%d/%y"),
  Sunday = 1:3,
  Monday = 4:6,
  Tuesday = 7:9,
  Wednesday = 10:12,
  Thursday = 13:15,
  Friday = 16:18,
  Saturday = 19:21
                  ))
```

You'll see that using the `gather` function as we did before makes the data frame out of order, which isn't very useful if we wanted to plot the data over time or do some other task that requires the data sorted by day. We can overcome this simply by using dplyr's `arrange` function. 
```{r}
daily <- dat %>%
  gather(key = Day, value = "x", Sunday:Saturday) %>%
  arrange(week_begins)

head(daily)
```

Suppose instead of gathering several variables into one variable, you wanted to do the opposite by take a single variable and spread it out into different variables. For example, 
suppose we wanted to the daily data set and create the dat data set by turning the variable x in the daily set into the variables Sunday, Monday, ... like in that dat data set. 

We can use the tidyr function `spread` to do this. As usual the first argument is the data set. The second argument is column whose values will you are transforming into column names in the output data set; in our case the "Day" column in the daily data set. The third argument is the column whose values will populate the cells; in this case the "x" column in the daily data set. 

We'll add one more line of code using the dplyr `select` function so we can order the columns as they are in the dat data set.
```{r}
daily %>%
  spread(Day, x) %>%
  select(week_begins, Sunday, Monday, Tuesday, Wednesday, Thursday, Friday, Saturday)
```

We can do this because each x in the daily  data set can be identified as belonging to a specific Day and week_begins value. However, if we tried to get back the iris data set from the iris2 data set because not every value of new_col can be matched to a specific species and measurement_type. For example, the first row in iris2 has a value for new_col of 5.1. We can deduce from iris2 that this is a sepal length from a setosa flower. Let's call this flower1. Now can we reliably deduce flower1's sepal width? From this data set not really. However, we could overcome this by inserting a flower id column in the original iris data set.

```{r}
iris3 <- iris %>%
  mutate(id = 1:nrow(iris))

iris4 <- iris3 %>%
  gather( key = "measurement_type", value = "new_col", Sepal.Length:Petal.Width)


head(iris4)
```
Now, we can transform iris4 back into the iris3 data set as follows.

```{r}
iris4 %>%
  spread(measurement_type, new_col)
```








